---
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
---

```{r intro, echo=FALSE, results="hide"}
knitr::opts_chunk$set(echo=FALSE, 
                      message=FALSE, 
                      comment = "", 
                      warning=FALSE, 
                      results="hide") 

date<-as.Date(Sys.time(	), format='%d%b%Y')
time<-Sys.time()

suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(readxl)))
suppressWarnings(suppressMessages(library(plotly)))
suppressWarnings(suppressMessages(library(Matrix)))
suppressWarnings(suppressMessages(library(stringr)))
suppressWarnings(suppressMessages(library(stringi)))
suppressWarnings(suppressMessages(library(readxl)))
suppressWarnings(suppressMessages(library(lubridate)))
suppressWarnings(suppressMessages(library(zoo)))
```

```{r dtaRegion}
#estimates:Total population, both sexes combined, as of 1 July (thousands)
dtaregion<-read_excel("~/Dropbox/0 Project/COVID19_US/DataUSDA/StatesByCensusRegionAndDivision.xlsx", sheet="Sheet1") %>%
    select(state, region, division)%>%
    filter(is.na(state)==F)  

str(dtaregion)
length(unique(dtaregion$state))
```

```{r dtaPop}
#estimates:Total population, both sexes combined, as of 1 July (thousands)
dtapop<-read.csv("~/Dropbox/0 Project/COVID19_US/DataCensusBureau/co-est2019-alldata.csv")

dtapop<-dtapop%>%
    mutate(state=as.character(STNAME), 
           pop=as.numeric(POPESTIMATE2019))%>%
    filter(COUNTY==0)%>%
    select(state, pop)

str(dtapop)
length(unique(dtapop$state))
```

```{r dtaCOVID}
url<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"

dtacases<-read.csv(url)%>%
    rename(country = Country_Region,
           state = Province_State,
           county = Admin2)%>%
    select(country, state, county, FIPS, Combined_Key, starts_with("X"))%>%
    gather(variable, value, starts_with("X"))%>%
    rename(date = variable,
           cases = value)%>%
    mutate(country=as.character(country),
           state=as.character(state),
           county=as.character(county))%>%
    select(country, state, date, cases)%>%
    group_by(country, state, date)%>%
    summarize_all(funs(sum))%>%ungroup()

url<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

dtadeaths<-read.csv(url)%>%
    rename(
           state = Province_State,
           county = Admin2)%>%
    select(state, county, FIPS, Combined_Key, starts_with("X"))%>%
    gather(variable, value, starts_with("X"))%>%
    rename(date = variable,
           deaths = value)%>%
    mutate(
           state=as.character(state),
           county=as.character(county))%>%
    select(state, date, deaths)%>%
    group_by(state, date)%>%
    summarize_all(funs(sum))%>%ungroup()

str(dtacases)
str(dtadeaths)

dtacovid<-left_join(dtacases, dtadeaths, by = c("state", "date"))%>%
    mutate(date=mdy(substring(date, 2)) )

dim(dtacases)
dim(dtadeaths)
dim(dtacovid)

```

```{r dta}
str(dtacovid)
str(dtapop)

dta<-full_join(dtacovid, dtapop, by = "state")

dim(dtacovid)
dim(dtapop)
dim(dta)

    temp<-dta%>%filter(is.na(pop)==TRUE) 
    table(temp$state) #territories without pop: DROP THESE below
    
    temp<-dta%>%filter(is.na(cases)==TRUE)
    table(temp$state) #territories/States without COVID data = 0

dta<-dta%>%filter(is.na(pop)==FALSE)%>%
    arrange(state, date)%>%
    mutate(
        newcases=cases-lag(cases),
        newcases=ifelse(state!=lag(state), NA, newcases),
        
        newdeaths=deaths-lag(deaths),
        newdeaths=ifelse(state!=lag(state), NA, newdeaths),
        
        incidence=round(100000*cases/pop, 3), # confiremd cases per 100,000 pop
        cfr=round(100*deaths/cases, 3), # deaths per 100 confirmed cases   
        mortality=round(100000*deaths/pop, 3) # deaths per 100000 pop
    )%>%arrange(state, date)%>%
    mutate(
        newcasessmooth =c(NA,NA,NA,NA,NA,NA,rollmean(newcases, 7)), 
        newdeathssmooth=c(NA,NA,NA,NA,NA,NA,rollmean(newdeaths, 7)), 
        newcasessmooth=ifelse(state!=lag(state), NA, newcasessmooth),
        newdeathssmooth=ifelse(state!=lag(state), NA, newdeathssmooth),
        newcasessmooth =round(newcasessmooth, 3),  
        newdeathssmooth=round(newdeathssmooth, 3),

        newcasessmoothpp=round(100000*newcasessmooth/pop, 1),   
        newdeathssmoothpp=round(100000*newdeathssmooth/pop, 1)

    )

#data quality problem: cumulative deaths and acses going down in some cases 
    
    summary(dta$newcases)
    summary(dta$newdeaths)
    temp<-dta%>%filter(newcases<0 | newdeaths<0)
    dim(temp)
    head(temp, nrow(temp))
    
    temp<-dta%>%filter(newcasessmooth<0 | newdeathssmooth<0)
    dim(temp)
    head(temp, nrow(temp))    

#Force negative smooth numbers to 0 : mainly for new deaths in 8 observations 

dta<-dta%>%
    mutate(
        newcasessmooth   =ifelse(newcasessmooth<0,    0, newcasessmooth),
        newcasessmoothpp =ifelse(newcasessmoothpp<0,  0, newcasessmoothpp),
        newdeathssmooth  =ifelse(newdeathssmooth<0,   0, newdeathssmooth),
        newdeathssmoothpp=ifelse(newdeathssmoothpp<0, 0, newdeathssmoothpp)
    )

dta<-dta%>%filter(date>="2020-03-01")

```

```{r dtaus}

dtaus<-dtacovid%>%
    select(date, cases, deaths)%>%
    group_by(date)%>%
    summarize_all(funs(sum))%>% #collapse cases and deaths 
    mutate(pop=sum(dtapop$pop))%>% #total popualtion 
    arrange(date)%>%
    mutate(
        newcases=cases-lag(cases),
        newdeaths=deaths-lag(deaths),

        newcasessmooth =c(NA,NA,NA,NA,NA,NA,rollmean(newcases, 7)), 
        newdeathssmooth=c(NA,NA,NA,NA,NA,NA,rollmean(newdeaths, 7)), 

        newcasessmooth =round(newcasessmooth, 3),  
        newdeathssmooth=round(newdeathssmooth, 3),

        newcasessmoothpp_US=round(100000*newcasessmooth/pop, 1),   
        newdeathssmoothpp_US=round(100000*newdeathssmooth/pop, 1))%>%
    filter(date>="2020-03-01")%>%
    select(date, newcasessmoothpp_US)

```

```{r dtacurve}

dtacurve<-left_join(dta, dtaregion, by = c("state"))

    dim(dta)
    dim(dtaus)
    dim(dtacurve)
    
dtacurve<-left_join(dtacurve, dtaus, by = c("date"))

    dim(dtacurve)
    
dtacurve<-dtacurve%>%
    arrange(country, state, date)%>%
    group_by(state)%>%
    mutate(
        day = row_number(),
        maxday = max(day),
        latestmonth= maxday-day<=30, 
        newcasessmoothpplatestmonth = newcasessmoothpp,
        newcasessmoothpplatestmonth = ifelse(latestmonth!=1, NA,
                                             newcasessmoothpplatestmonth), 
        
        latest=date==max(date), 
        
        peakcasespp=round(max(newcasessmoothpp, na.rm = TRUE), 1), 
        peakdate=as.character(date),
            peakdate=ifelse(peakcasespp!=newcasessmoothpp, "", peakdate),
            peakdate=ymd(substring(peakdate, 1)) )%>%
    fill(peakdate)%>%
    fill(peakdate, .direction = "up")%>%
    mutate(
        latestcasespp=newcasessmoothpp, 
        latestcasespp=ifelse(latest==FALSE, NA, latestcasespp),
        latestlevel=round((latestcasespp)/(peakcasespp), 3)
        )%>%
    fill(latestlevel)%>%
    fill(latestlevel, .direction = "up")%>%
    ungroup()%>%
    select(country, region, division, state, date, cases, deaths, pop, incidence, cfr, mortality, starts_with("new"), starts_with("start"), starts_with("peak"), starts_with("latest"))


```

### __COVID-19 waves in the US: How is my state doing? Seeing is believing.__  

<div class="alert alert-info">
__NOTE__   
Since July 2, 2022, select data from this site have been presented in __[an interactive app](https://isquared.shinyapps.io/COVID19_US_States)__. This site is not maintained regularly anymore.
</div>

(Last updated on `r date`. Published initially on November 1, 2020.) 

[_For further data, including_ __states ranked by the latest 7-day average of new cases per population,__ _please see this._](https://rpubs.com/YJ_Choi/COVID19_US_supplement)

[_For the earlier publication,_ __"Is my state ready to reopen? Seeing is believing,"__ _please see this._](https://rpubs.com/YJ_Choi/COVID19_US_states_version1)

The first COVID-19 case in the US was reported on January 3rd 2020. Since March, many states have implemented various strategies to "flatten the curve". In late Spring when most states were in their first wave, the most dominant question was [if our states were ready to reopen](https://medium.com/@yj.choi.isquared/is-my-state-ready-to-reopen-seeing-is-believing-41ef39de663a).  

Since then states have experienced vastly different curves to date. Some are in the middle of the first wave - which can be unprecedentedly high, unfortunately. A majority of states, however, now face new waves of COVID-19 - although distinguishing a wave from another is not always straight forward or meaningful. 

We follow each state's curve below, organized by geography. States are grouped by [USDA's geographic region and division classification](https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/docs/regions/).   

```{r plotpanelRegion}    
mindate<-min(dtacurve$date)
maxdate<-max(dtacurve$date)

panel <- . %>% 
    plot_ly(x = ~date)%>% 
    add_trace(
        y = 0, type = 'scatter', mode = 'lines',
        line= list(color = "#F1F1F1") ) %>%    
    add_trace(
        y = 10, type = 'scatter', mode = 'lines',
        fill = 'tonexty',fillcolor='#EEEEEE', opacity=0.1,
        line= list(color = "#F1F1F1") ) %>%    
    add_lines(y = ~newcasessmoothpp_US,
        line= list(color = "#BEBEBE"),
        hoverinfo = 'text',
        text = ~paste(
                      '</br> United States',          
                      '</br> Date: ', date,
                      '</br> New cases per 100,000: ', newcasessmoothpp_US)  )%>%
    add_lines(y = ~newcasessmoothpp,
        line= list(color = "#70A8CF"),  
        hoverinfo = 'text',
        text = ~paste(
                      '</br> State: ', state,          
                      '</br> Date: ', date,
                      '</br> New cases per 100,000: ', newcasessmoothpp)  )%>%
    add_trace(
        y = ~newcasessmoothpplatestmonth, type = 'scatter', mode = 'lines',
        line= list(color = "#195F90") ) %>%
    add_trace(
        y = ~latestcasespp, type = 'scatter', mode = 'markers',
        marker = list(size = 5,color ="#70A8CF"),
        text = ~latestcasespp, 
        textfont=list(size=10, color="#70A8CF"), 
        textposition = "bottom left"
        )%>%  
    add_annotations(
        text = ~unique(state),
        x = 0.5, y = 0.90, xref = "paper", yref = "paper",    
        xanchor = "center", yanchor = "bottom", showarrow = FALSE,
        font = list(size = 12)  )%>%
    layout(
        showlegend = FALSE,
        yaxis=list(title="New cases per 100,000" , 
                   range=c(0,maxy), showgrid = FALSE),
        xaxis=list(title="" , range=c(mindate,maxdate), showgrid = FALSE)
        ) 

```

---

#### __Northeast__
_Note: The x-axis is date since March. The y-axis is 7-day rolling average of daily new cases per 100,000 population. <span style="color: #195F90;">The darker blue section of the line represents the lastest 30 days.</span> The gray line is the US national data. The light gray box represents the number of daily new cases 10 or lower per 100,000._  

_Hover over the figures to see the underlying data._
```{r plotNortheast, results="asis", fig.align="left", out.width="800px", out.height="650px"}

dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="Northeast")
maxy<-max(dtafig$newcasessmoothpp)
nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

---

#### __South Atlantic__
_Note: The x-axis is date since March. The y-axis is 7-day rolling average of daily new cases per 100,000 population. <span style="color: #195F90;">The darker blue section of the line represents the lastest 30 days.</span> The gray line is the US national data. The light gray box represents the number of daily new cases 10 or lower per 100,000._  

_Hover over the figures to see the underlying data._  
```{r plotSouthAtlantic, results="asis", fig.align="left", out.width="800px", out.height="650px"}

dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="South" & division=="South Atlantic")
maxy<-max(dtafig$newcasessmoothpp)

nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

---

#### __South Central__
_Note: The x-axis is date since March. The y-axis is 7-day rolling average of daily new cases per 100,000 population. <span style="color: #195F90;">The darker blue section of the line represents the lastest 30 days.</span> The gray line is the US national data. The light gray box represents the number of daily new cases 10 or lower per 100,000._  

_Hover over the figures to see the underlying data._
```{r plotSouthCentral, results="asis", fig.align="left", out.width="800px", out.height="450px"}

dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="South" & division!="South Atlantic")
maxy<-max(dtafig$newcasessmoothpp)

nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

---

#### __Midwest__
_Note: The x-axis is date since March. The y-axis is 7-day rolling average of daily new cases per 100,000 population. <span style="color: #195F90;">The darker blue section of the line represents the lastest 30 days.</span> The gray line is the US national data. The light gray box represents the number of daily new cases 10 or lower per 100,000._  

_Hover over the figures to see the underlying data._
```{r plotMidwest, results="asis", fig.align="left", out.width="800px", out.height="250px", eval=FALSE}

dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="Midwest" & peakcasespp<100)
maxy<-max(dtafig$newcasessmoothpp)

nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

```{r plotMidwestHigh, results="asis", fig.align="left", out.width="800px", out.height="450px"}
dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="Midwest" & peakcasespp>=100)
maxy<-max(dtafig$newcasessmoothpp)

nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

---

#### __West__
_Note: The x-axis is date since March. The y-axis is 7-day rolling average of daily new cases per 100,000 population. <span style="color: #195F90;">The darker blue section of the line represents the lastest 30 days.</span> The gray line is the US national data. The light gray box represents the number of daily new cases 10 or lower per 100,000._  

_Hover over the figures to see the underlying data._
```{r plotWest, results="asis", fig.align="left", out.width="800px", out.height="1050px"}

dtafig<-dtacurve%>%
    mutate(latestcasespp=round(ifelse(latest==FALSE, NA, latestcasespp), 0))%>%
    filter(region=="West")
maxy<-max(dtafig$newcasessmoothpp)

nobs<-ceiling(length(unique(dtafig$state))/4)    

dtafig%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = nobs, shareX = TRUE, shareY = TRUE)  
```

---

All COVID-19 data (i.e., cumulative confirmed cases and deaths by day) come from [JHU/CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data), accessed on `r date`. All data on state population come from [US Census Bureau](https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-total.html#par_textimage), accessed on March 29, 2020.   

For those interested in the methods and more data by state, [please see this](https://rpubs.com/YJ_Choi/COVID19_US_supplement).   

For those interested in the US curve, compared to other countries, [please see this](https://medium.com/@yj.choi.isquared/flattening-covid-19-curve-in-south-korea-and-comparison-among-oecd-countries-singapore-and-taiwan-ae211a5645c9). 

---

<p style="color:gray">
See [GitHub](https://github.com/yoonjoung/COVID19_FlattenedCurve) for data, code, and more information. 
For typos, errors, and questions, contact me at [www.isquared.global](https://www.iSquared.global/YJ). 

_Making Data Delicious, One Byte at a Time_, in good times and bad times.</p>