---
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
---
```{r intro, echo=FALSE, results="hide"}
knitr::opts_chunk$set(echo=FALSE, 
                      message=FALSE, 
                      comment = "", 
                      warning=FALSE, 
                      results="hide") 

date<-as.Date(Sys.time(	), format='%d%b%Y')
time<-Sys.time()

suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(readxl)))
suppressMessages(library(plotly))
suppressMessages(library(Matrix))
suppressMessages(library(stringr))
suppressMessages(library(stringi))
suppressWarnings(suppressMessages(library(readxl)))
suppressWarnings(suppressMessages(library(lubridate)))
suppressWarnings(suppressMessages(library(zoo)))
```

```{r dtaPop}
#estimates:Total population, both sexes combined, as of 1 July (thousands)
dtapop<-read.csv("C:/Users/YoonJoung Choi/Dropbox/0 Project/COVID19_US/DataCensusBureau/co-est2019-alldata.csv")

dtapop<-dtapop%>%
    mutate(state=as.character(STNAME), 
           pop=as.numeric(POPESTIMATE2019))%>%
    filter(COUNTY==0)%>%
    select(state, pop)

str(dtapop)
length(unique(dtapop$state))
```


```{r dtaCOVID}
url<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"

dtacases<-read.csv(url)%>%
    rename(country = Country_Region,
           state = Province_State,
           county = Admin2)%>%
    select(country, state, county, FIPS, Combined_Key, starts_with("X"))%>%
    gather(variable, value, starts_with("X"))%>%
    rename(date = variable,
           cases = value)%>%
    mutate(country=as.character(country),
           state=as.character(state),
           county=as.character(county))%>%
    select(country, state, date, cases)%>%
    group_by(country, state, date)%>%
    summarize_all(funs(sum))%>%ungroup()


url<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

dtadeaths<-read.csv(url)%>%
    rename(
           state = Province_State,
           county = Admin2)%>%
    select(state, county, FIPS, Combined_Key, starts_with("X"))%>%
    gather(variable, value, starts_with("X"))%>%
    rename(date = variable,
           deaths = value)%>%
    mutate(
           state=as.character(state),
           county=as.character(county))%>%
    select(state, date, deaths)%>%
    group_by(state, date)%>%
    summarize_all(funs(sum))%>%ungroup()

str(dtacases)
str(dtadeaths)

dtacovid<-left_join(dtacases, dtadeaths, by = c("state", "date"))%>%
    mutate(date=mdy(substring(date, 2)) )

dim(dtacases)
dim(dtadeaths)
dim(dtacovid)
```

```{r dta}
str(dtacovid)
str(dtapop)

dta<-full_join(dtacovid, dtapop, by = "state")

dim(dtacovid)
dim(dtapop)
dim(dta)

    temp<-dta%>%filter(is.na(pop)==TRUE) 
    table(temp$state) #territories without pop: DROP THESE below
    
    temp<-dta%>%filter(is.na(cases)==TRUE)
    table(temp$state) #territories/States without COVID data = 0

dta<-dta%>%filter(is.na(pop)==FALSE)%>%
    arrange(state, date)%>%
    mutate(
        newcases=cases-lag(cases),
        newcases=ifelse(state!=lag(state), NA, newcases),
        
        newdeaths=deaths-lag(deaths),
        newdeaths=ifelse(state!=lag(state), NA, newdeaths),
        
        incidence=round(100000*cases/pop, 1), # confiremd cases per 100,000 pop
        cfr=round(100*deaths/cases, 1), # deaths per 100 confirmed cases   
        mortality=round(100000*deaths/pop, 1) # deaths per 100000 pop
    )%>%arrange(state, date)%>%
    mutate(
        newcasessmooth =c(NA,NA,NA,rollmean(newcases, 7),NA,NA,NA), 
        newdeathssmooth=c(NA,NA,NA,rollmean(newdeaths, 7),NA,NA,NA), 
        newcasessmooth=ifelse(state!=lag(state), NA, newcasessmooth),
        newdeathssmooth=ifelse(state!=lag(state), NA, newdeathssmooth),

        newcasessmoothpp=round(100000*newcasessmooth/pop, 1),   
        newdeathssmoothpp=round(100000*newdeathssmooth/pop, 1)

    )

#data quality problem: cumulative deaths and acses going down in some cases 
    
    summary(dta$newcases)
    summary(dta$newdeaths)
    temp<-dta%>%filter(newcases<0 | newdeaths<0)
    dim(temp)
    head(temp, nrow(temp))
    
    temp<-dta%>%filter(newcasessmooth<0 | newdeathssmooth<0)
    dim(temp)
    head(temp, nrow(temp))    

#Force negative smooth numbers to 0 

dta<-dta%>%
    mutate(
        newcasessmooth   =ifelse(newcasessmooth<0,    0, newcasessmooth),
        newcasessmoothpp =ifelse(newcasessmoothpp<0,  0, newcasessmoothpp),
        newdeathssmooth  =ifelse(newdeathssmooth<0,   0, newdeathssmooth),
        newdeathssmoothpp=ifelse(newdeathssmoothpp<0, 0, newdeathssmoothpp)
    )
```

```{r dtacurve}

dtacurve<-dta%>%filter(incidence>1)
    
    length(unique(dtacurve$state))
    length(unique(dta$state))    

dtacurve<-dtacurve%>%
    arrange(country, state, date)%>%
    group_by(state)%>%
    mutate(
        day = row_number(),
        latest=date==max(date),
        
        startdate=min(date),
        startcases=newcasessmooth,
            startcases=ifelse(date!=startdate, NA, startcases), 
        startcasespp=round(100000*startcases/pop, 1))%>%
    fill(startcases, startcasespp)%>%
    fill(startcases, startcasespp, .direction = "up")%>%
    mutate(
        peakcases=max(newcasessmooth, na.rm = TRUE),
        peakcasespp=round(100000*peakcases/pop, 1),
        peakdate=as.character(date),
            peakdate=ifelse(peakcases!=newcasessmooth, "", peakdate),
            peakdate=ymd(substring(peakdate, 1)) )%>%
    fill(peakdate)%>%
    fill(peakdate, .direction = "up")%>%
    mutate(
        endcases=newcasessmooth,
            endcases=ifelse(date<=peakdate, NA, endcases), 
            endcases=ifelse(newcasessmooth>startcases, NA, endcases),
        enddate=as.character(date),
            enddate=ifelse(is.na(endcases)==TRUE, "", enddate),
            enddate=ifelse(is.na(lag(endcases))==FALSE & is.na(endcases)==FALSE, "", enddate),
            enddate=ymd(substring(enddate, 1)))%>%
    fill(enddate)%>%
    fill(enddate, .direction = "up")%>%
    ungroup()%>%
    select(country, state, date, day, latest, cases, deaths, pop,  incidence, cfr, mortality, starts_with("new"), starts_with("start"), starts_with("peak"), starts_with("end"), -endcases)


```


####__Flattening COVID-19 curve in the US: state-level data__  

(Last updated on `r date`)

This replicates [comparison of the epidemic curve among South Korea, Singapore, Taiwan, and OECD country comparison analysis](https://rpubs.com/YJ_Choi/KCOVID), using US state-level data. 

_Hover over each figure to see values and more options_. States are color-coded by [USDA's geogrpahic region](https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/docs/regions/).   

_See data sources and methods at the end_.  

####__1. Latest snapshot by state__

__Cumulative incidence rate__: number of confirmed cases per 100,000 population. It may not be a good indicator to compare, considering vastly different testing rates - especially in the beginning of the epidemic. Also, definitions of confirmed cases can be slightly different across states. 

```{r}
#color by USDA region and division
colorlist<- list(color = c(
 '#fdae61' ,
 '#a6d96a' ,
 '#a6d96a' ,
 '#fee08b' ,
 '#a6d96a' ,
 '#a6d96a' ,
 '#fdae61' ,
 '#fee08b' ,
 '#fee08b' ,
 '#fee08b' ,
 '#fee08b' ,
 '#a6d96a' ,
 '#a6d96a' ,
 '#d9ef8b' ,
 '#d9ef8b' ,
 '#d9ef8b' ,
 '#d9ef8b' ,
 '#fdae61' ,
 '#fee08b' ,
 '#fdae61' ,
 '#fee08b' ,
 '#fdae61' ,
 '#d9ef8b' ,
 '#d9ef8b' ,
 '#fdae61' ,
 '#d9ef8b' ,
 '#a6d96a' ,
 '#d9ef8b' ,
 '#a6d96a' ,
 '#fdae61' ,
 '#fdae61' ,
 '#a6d96a' ,
 '#fdae61' ,
 '#fee08b' ,
 '#d9ef8b' ,
 '#d9ef8b' ,
 '#fee08b' ,
 '#a6d96a' ,
 '#fdae61' ,
 '#fdae61' ,
 '#fee08b' ,
 '#d9ef8b' ,
 '#fdae61' ,
 '#fee08b' ,
 '#a6d96a' ,
 '#fdae61' ,
 '#fee08b' ,
 '#a6d96a' ,
 '#fee08b' ,
 '#d9ef8b' ,
 '#a6d96a' ))
```

    
```{r plotincidence, results="asis", fig.align="left", out.width="800px"}

dtafig<-dtacurve%>%filter(latest==TRUE)    

dtafig$state<-factor(dtafig$state, 
                    levels = unique(dtafig$state) 
                    [order(dtafig$incidence, decreasing = TRUE)])


plot_ly(dtafig, x=~state, y=~incidence, type = 'bar', marker=colorlist)%>%
    layout(
            title = c("COVID-19 confirmed cases per 100,000 population"),     
            xaxis = list(title = "State",  
                         autotick = FALSE,
                         showticklabels = TRUE, 
                         tickfont = list(size=8), 
                         tickangle=-90),
            yaxis = list(title = "per 100,000 population") 
            )
```

__Mortality rate__: number of COVID-19 deaths per 100,000 population. Considering different testing rates and this may be the most appropriate indicator to compare impact of the epidemic across states.  

```{r plotmortality, results="asis", fig.align="left", out.width="800px"}

dtafig<-dtacurve%>%filter(latest==TRUE)     

dtafig$state<-factor(dtafig$state, 
                    levels = unique(dtafig$state) 
                    [order(dtafig$mortality, decreasing = TRUE)])

plot_ly(dtafig, x=~state, y=~mortality, type = 'bar', marker=colorlist,
              name= "State")%>%
    layout(
            title = c("COVID-19 deaths per 100,000 population"),     
            xaxis = list(title = "State",  
                         autotick = FALSE,
                         showticklabels = TRUE, 
                         tickfont = list(size=8), 
                         tickangle=-90),
            yaxis = list(title = "per 100,000 population") 
            )

```

And, for those who wonder about __Case fatality rate__ (CFR): number of deaths per 100 confirmed cases. CFR is sensitive to demographic and clinical characteristics of patients. CFR can be high in states where testing has been limited to only people with clinical symptoms. 

```{r plotcfr, results="asis", fig.align="left", out.width="800px"}

dtafig<-dtacurve%>%filter(latest==TRUE)    

dtafig$state<-factor(dtafig$state, 
                    levels = unique(dtafig$state) 
                    [order(dtafig$cfr, decreasing = TRUE)])

plot_ly(dtafig, x=~state, y=~cfr, type = 'bar', marker=colorlist,
              name= "State")%>%
    layout(
            title = c("COVID-19 deaths per 100 confirmed cases (Case Fatality Rate)"),     
            xaxis = list(title = "State",  
                         autotick = FALSE,
                         showticklabels = TRUE, 
                         tickfont = list(size=8), 
                         tickangle=-90),
            yaxis = list(title = "deaths per 100 confirmed cases") 
            )
```

####__2. Epidemic curve by state__

The epidemic curve in this analysis was constructed based on __the number of new confirmed cases each day__. Considering vastly different population sizes, _the number of new confirmed cases each day per 100,000 population_ was used, instead of the absolute number. In addition, _7-day rolling averages_ were used to avoid any isolated peaks/drops, which can be caused by various reasons other than the true course of the epidemic itself (e.g., changes in definition, and lab process delay). 

Let's focus on the __first wave__ of the epidemic, which we define to start when cumulative incidence exceeds 1 confirmed case per 100,000 population. The __peak__ is when the number of new confirmed cases is at its maximum. There are three distinctive phases:  

* Pre peak (accelerating growth phase - _red shade_ in the below figure): Period when the number of new confirmed cases increases. In many states (though not all), we have seen an exponential growth in the total confirmed cases during this phase.    
* Post peak (decelerating growth phase - _orange shade_): Period when the number of new confirmed cases declines, to the level/threshold where the exponential growth started.   
* Stable to minimal growth (_yellow shade_): the number of new confirmed cases remain more or less stable - under the threshold. Incidence still increases, but at a much slower rate than before. __This last phase provides critical opportunities to realign resources in order to  prevent, delay, and flatten the next wave__ in a more proactive way than reacting to crises. 

With successful control of the epidemic, we will see both narrow and also short peak.  

#####__2.1 Vermont, as an example__
```{r}

temp<-dtacurve%>%filter(state=="Vermont") 

startdate<-max(temp$startdate, na.rm = TRUE)
peakdate<-max(temp$peakdate, na.rm = TRUE)
peakcasespp<-max(temp$peakcasespp, na.rm = TRUE)
```

The data (i.e., smoothed number of new confirmed cases each day per 100,000 population) show that Vermont has well passed the peak, and __is approaching the stable phase as long as the pace continues__. The cumulative incidence rate exceeded 1 per 100,000 population on `r startdate`, and the daily new cases reached the peak (`r peakcasespp` new confirmed cases per 100,000 population) on `r peakdate`.  

```{r plotVT, results="asis", fig.align="left", out.width="800px"}
dtafig<-dtacurve%>%filter(state=="Vermont")

shapelist<-list(
            list(
                type = "rect",  fillcolor = "#f03b20", 
                line = list(color = "#f03b20"), opacity = 0.2,
                x0=min(dtafig$date), x1=max(dtafig$peakdate, na.rm = TRUE),xref = "x",
                y0=0,y1=min(dtafig$peakcasespp, na.rm = TRUE), yref = "y2") , 
            list(
                type = "rect",  fillcolor = "#fd8d3c", 
                line = list(color = "#fd8d3c"), opacity = 0.2,
                x0=min(dtafig$peakdate, na.rm = TRUE), x1=max(dtafig$enddate, na.rm = TRUE),xref = "x",
                y0=0,y1=min(dtafig$peakcasespp, na.rm = TRUE), yref = "y2" ),
            list(
                type = "rect",  fillcolor = "#fed976", 
                line = list(color = "#fed976"), opacity = 0.2,
                x0=max(dtafig$enddate, na.rm = TRUE),x1=max(dtafig$date),xref = "x",
                y0=0,y1=min(dtafig$peakcasespp, na.rm = TRUE), yref = "y2" )
            )

plot_ly(dtafig, x=~date, y=~incidence, name="Total cases", 
        type='bar', marker=list(color = c( "lightgray"))) %>%
    add_trace(y=~newcasessmoothpp, name="New cases",
        type='scatter',mode = 'lines', 
        marker=list(color = c( "black"), size = 1),
        line=list(color = c( "black")), yaxis='y2')%>%
    layout(
        
        title = c("Trend of daily new confirmed cases per 100,000 population since the first wave"),
        xaxis = list(title = "", tickfont = list(size=10)), 
        yaxis2 = list(overlaying='y',side="right",
                      title = "Daily new cases per 100,000 population"),
        yaxis = list(title = "Total cases per 100,000 population"),
        margin = list(b = 100, r=100), 
        legend=list(orientation="h", x = 0.3, y = -0.1)
        )

```

#####__2.2 Curve by state__
```{r}
temp<-dtacurve%>%filter(is.na(enddate)==FALSE)
table(temp$country)
nstable<-length(unique(temp$state))

```

_Holding beginning of the first wave on day 0_, the following figure compares the length and height of the peak across the states. 

* Only `r nstable` states, Montana and Hawaii, have entered the stable phase. 

* However, some would reach the stable phase soon if the pace continues (e.g., Vermont).  

* Still, most states passed the peak only recently. In these states, the peak can move up since fluctuating increases were observed in other states/countries.  

* Some have entered the first wave, and the curve is going up slowly.

States are presented in groups, due to varying range of peak height. Top panel has states with relatively lower peak, and the bottom panel shows states with higher peak.  

```{r plotSTATES1, results="asis", fig.align="left", out.width="800px", out.height="400px"}

panel <- . %>% 
  plot_ly(x = ~day, y = ~newcasessmoothpp) %>%
  add_lines(line=list(color=("'#1F77B4'"))) %>%
  add_annotations(
    text = ~unique(state),
    x = 0.5, y = 0.9, xref = "paper", yref = "paper",    
    yanchor = "bottom", showarrow = FALSE,
    font = list(size = 8)
  )%>%
  layout(
    showlegend = FALSE,
    yaxis=list(title="", range=c(0,8)),
    xaxis=list(title="", range=c(0,45))) 

dtacurve%>%filter(peakcasespp<median(peakcasespp))%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = 4, shareX = TRUE, shareY = TRUE)  
```

```{r plotSTATES2, results="asis", fig.align="left", out.width="800px", out.height="300px"}

panel <- . %>% 
  plot_ly(x = ~day, y = ~newcasessmoothpp) %>%
  add_lines(line=list(color=("'#1F77B4'"))) %>%
  add_annotations(
    text = ~unique(state),
    x = 0.5, y = 0.9, xref = "paper", yref = "paper",    
    yanchor = "bottom", showarrow = FALSE,
    font = list(size = 8)
  )%>%
  layout(
    showlegend = FALSE,
    title = c(""),
    yaxis=list(title="", range=c(0,25)),
    xaxis=list(title="", range=c(0,45))) 

dtacurve%>%filter(peakcasespp>=median(peakcasespp)& peakcasespp<25)%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = 3, shareX = TRUE, shareY = TRUE)  
```

```{r plotSTATES3, results="asis", fig.align="left", out.width="800px", out.height="200px"}

panel <- . %>% 
  plot_ly(x = ~day, y = ~newcasessmoothpp) %>%
  add_lines(line=list(color=("'#1F77B4'"))) %>%
  add_annotations(
    text = ~unique(state),
    x = 0.5, y = 0.9, xref = "paper", yref = "paper",    
    yanchor = "bottom", showarrow = FALSE,
    font = list(size = 8)
  )%>%
  layout(
    showlegend = FALSE,
    title = c(""),
    yaxis=list(title=""),
    xaxis=list(title="", range=c(0,45))) 

dtacurve%>%filter(peakcasespp>=median(peakcasespp) & peakcasespp>=25)%>%
    group_by(state) %>%
    do(p = panel(.)) %>%
    subplot(nrows = 1, shareX = TRUE, shareY = TRUE) 
```


####__3. Putting it all together: timeline, length and height of the peak__

To summarize, below first shows the timeline of the first wave and its phases.  
* The peak length is still to be determined in many states (with no yellow dot below).      

```{r plottimeline, results="asis", fig.align="left", out.width="800px", out.height="1000px"  }

dtafig<-dtacurve%>%filter(latest==TRUE) 

dtafig$state <- factor(dtafig$state, 
                         levels = unique(dtafig$state)
                         [order(dtafig$startdate, decreasing = TRUE)])


plot_ly(dtafig, color = I("gray80"))%>%
    add_segments(x = ~startdate, xend = ~peakdate, y = ~state, yend = ~state, showlegend = FALSE)%>% 
    add_segments(x = ~peakdate, xend = ~enddate, y = ~state, yend = ~state, showlegend = FALSE)%>% 
    add_markers(x = ~startdate, y = ~state, name = "Pre-peak phase starts", color = I("#f03b20"))%>% 
    add_markers(x = ~peakdate, y = ~state, name = "Post-peak phase starts", color = I("#fd8d3c"))%>% 
    add_markers(x = ~enddate, y = ~state, name = "Stable phase starts", color = I("#fed976") )%>%
layout(
    title="Timeline of the first wave of COVID-19", 
    yaxis = list(title=""),  
    xaxis = list(title="Date",tickfont = list(size=10)),
    legend = list(orientation = 'v',legendfont = list(size=8)))

```

The following compares the height of the peak (the maximum number of new confirmed cases per 100,000 population) by country.      

```{r plotheight, results="asis", fig.align="left", out.width="800px" }

dtafig<-dtacurve%>%filter(latest==TRUE) 

dtafig$state <- factor(dtafig$state, 
                         levels = unique(dtafig$state)
                         [order(dtafig$peakcasespp, decreasing = TRUE)])

plot_ly(dtafig, x=~state, y=~peakcasespp, 
        type = 'bar', marker=colorlist)%>%
    layout(
    yaxis = list(title="Peak number of new cases per 100,000 population"),  
    xaxis = list(title="",tickfont = list(size=10))  )

```

And, finally, if we put together the height of the peak (on a log scale, Y axis) and time to the peak (distance between the red and orange dots above)...       

```{r plotheightlength, results="asis", fig.align="left", out.width="800px"}

dtafig<-dtacurve%>%filter(latest==TRUE)%>%
    mutate(length=peakdate-startdate)

textlist <- list(
  size = 8,
  color = toRGB("grey50"))

plot_ly(dtafig, x=~length, y=~peakcasespp, 
        type = 'scatter', marker=colorlist,  
        text = ~state)%>%
    add_text(textfont = textlist, textposition = "top")%>%
    layout(
    showlegend=FALSE, 
    title="Length* and height of the peak",    
    yaxis = list(title="Peak number of new cases per 100,000 population", type = "log"),  
    xaxis = list(title="*Number of days between start to peak",tickfont = list(size=10))  )

```

---

__METHODS__  

__Data__:  
1. All COVID-19 data (i.e., cumulative confirmed cases and deaths by day) come from [JHU/CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data). Accessed on `r date`.    
2. All data on country population come from UN [World Population Prospects 2019 Revision](https://population.un.org/wpp/). Accessed on April 18, 2020.   

__Measures__:  
The number of new confirmed cases on each date was calculated based on the difference between cumulative numbers over two days. Then, a 7-day rolling average was calculated for, hereinafter referred to as the _smoothed_ number of new confirmed cases. Then, the smoothed number was divided by the total population in the country, as the __smoothed number of new confiremd cases per 100,000 popualtion__.   
* The first wave was defined to start when the cumulative incidence rate exceeds 1 per 100,000 population. The smoothed number of new confirmed cases at the start of the first wave varies by country, but it averages around 0.5 per 100,000.   
* A peak date of the wave was the date when the _smoothed number of new confiremd cases per 100,000 popualtion_ was the maximum.   
* A date entering the stable phase was when the _smoothed number of new confiremd cases per 100,000 popualtion_ was less than the number on the start date of the wave. 

COVID-mortality (either CFR or preferably COVID-specific mortality rate in the population) may be more comparable indicator, given considerably different testing strategies and testing rates. However, states are currently at different stages of the curve, and comparison of mortality data would be more appropriate once most states are in a similar phase of the epidemic (i.e., well in to the stable phase). 

---

<p style="color:gray">
See [GitHub](https://github.com/yoonjoung/COVID19_FlattenedCurve) for data, code, and more information. 
For typos, errors, and questions, contact me at [www.isquared.global]("https://www.iSquared.global/YJ")</p>

_Making Data Delicious, One Byte at a Time_, in good times and bad times.